\textbf{Transparency to hybrid open access through publisher-provided metadata: An article-level study of Elsevier}

### 2021-05-19 analysis/manuscript.Rmd
sv tmp

The rise of open access (OA) to journal articles has added complexity to scholarly publishing, particularly concerning transparency of economic dimensions. Financial transparency in journal publishing has long been lacking since Big Deal subscription contracts between academic institutions and large academic publishers usually include confidentiality clauses that prohibit publicly sharing information about the scope and pricing of such agreements [@Frazier_2001; @Bergstrom_2014; @Larivi_re_2015]. Further, the distributed process of paying for accessing and publishing scholarly articles, which can involve several actors at the same time, including authors, research organizations, libraries, and research funders, adds to the opacity of the financial flows of scholarly publishing [@Lawson_2016].

The problem with this lack of transparency becomes more apparent in the case of hybrid OA. In this OA business model, individual articles can be made openly available for a fee, also known as article-processing charge (APC), while the journal as a whole remains subscription access. Hence, hybrid OA was originally introduced as a low-risk transitional model that allows journals to gradually convert to full OA and reduce subscription costs as the OA uptake increases [@Prosser_2003]. Many large subscription publishers have since incorporated this OA model, and while the uptake among authors increased only slowly during the first decade [@Bj_rk_2012], it has gradually gained popularity in recent years. This development was driven by the implementation of OA policies by research institutions and funders, the allocation of OA funds, and formal agreements between publishers and institutions that allow affiliated authors to publish OA free-of-charge [@Laakso_2016]. However, there have been concerns that the hybrid model allows for double-dipping—receiving two payments for one article, the APC and subscription fees [@Shieber_2009; @Prosser_2015]. While publishers assure they combine subscription fees and APCs to adjust pricing [@Mittermaier_2015], without transparency around the uptake of hybrid OA and revenue streams, such claims are impossible to evaluate.

With the recent introduction of transformative agreements, an evolving term for contracts that shift library spendings for subscriptions to OA [@Borrego_2020; @Hinchliffe_2019], the demand for publisher-provided data has even increased. If publishers would enable transparency about the OA uptake in hybrid journals and APC invoicing, academic libraries could improve their assessments and adjustments of contracts with publishers [@antelmann_2017; @Schimmer_2015]. Likewise, research organizations and funders could improve compliance monitoring with OA mandates, while ensuring that the different actors were not charged multiple times [@Larivi_re_2018]. However, as formal bibliometrics and financial studies have noted, the lack of publicly available and standardized data about hybrid OA publishing is considerable [@Laakso_2016; @Lawson_2015; @Pinfield_2016]

This current study investigates the extent to which publisher-provided data can be utilized to research hybrid OA. The study takes Elsevier, the largest scholarly journal publisher in terms of article volume [@Larivi_re_2015], as the starting point for an empirical analysis of hybrid OA uptake and invoicing. Our approach is based on openly available data sources and brings together scholarly metadata from Crossref, a DOI registration agency for scholarly works, and machine-readable APC invoicing data from Elsevier OA full-texts. The so-compiled data enable us to analyze the development of Elsevier’s hybrid journal portfolio from 2015-19, and to compare it with that of Elsevier’s full and delayed OA program. Further, we investigate whether hybrid publishing fees were waived or charged. In particular, we identify academic consortia and research funders that were directly invoiced by Elsevier, and compare their hybrid OA rates by license and subject. As such, our findings have implications for research and OA policy and highlight the potential of publisher-provided data for large-scale, instantaneous analysis of hybrid OA uptake and invoicing.

@@

The ability to identify and discover hybrid OA articles plays a fundamental role in monitoring hybrid OA, yet time and again, previous studies have underscored the difficulties involved in this (cf. [@Akbaritabar_2019; @van_Leeuwen_2018; @Laakso_2016; @Emery_2013; @Bj_rk_2012]. Insights into the volume of hybrid OA and the various levels and sources of APC funding could create a better understanding of the costs involved in scholarly publishing and, thus, strengthen institutions' bargaining power in negotiations with publishers, and promote greater equity in scholarly publishing. In the following section, we examine how previous research has tackled the challenge of identifying hybrid OA content and what these studies taught us about the uptake and funding sources of hybrid OA.

@@

In the first article-level study on hybrid OA, @Laakso_2016 reported that between 2007 and 2013, the five largest publishers--Elsevier, Sage, Springer, Taylor & Francis, and Wiley--recorded growing numbers of hybrid journals that coincided with a twenty-fold increase in hybrid OA articles (from 666 to 13,994). As a result, hybrid journals with at least one OA article more than doubled in number (1,082 in 2009, 2,714 in 2013) but fell by 31% relative to the number of hybrid journals in total. Although Laakso and Björk's (2016) presented a comprehensive overview tracking the growth of hybrid OA over time and across multiple publishers, their exploratory approach is unsuitable for repeated use as it required a considerable amount of manual data collection and cleaning. Minimizing such manual tasks, @Nelson_2017 narrowed their focus to one publisher--the American Chemical Society (ACS)--and requested a list of all hybrid OA articles published from 2006-2011 directly from the ACS. To assess the hybrid OA uptake, the authors aggregated and compared the number of OA (n=814) and subscription articles (n=27,621), finding that over the five years, 2.9% of ACS articles were published as hybrid OA.

Contrary to the previous two studies, @Kirkman_2018 took a research funder as a starting point, reporting a 26% share of hybrid OA for articles funded by the Australian National Health and Medical Research Council from 2012-2014 (816 of 3,190). Kirkman collected research articles using the Funding Agency and Funding Text search fields on Web of Science, and classified articles as OA when they were freely available on the publisher website, and further distinguished hybrid OA articles based on journal-level information from the publisher website, such as hybrid OA journal and APC lists. Another approach is presented by @P_l_nen_2020, who used current research information system (CRIS) data—institutional publication data—to study the extent of OA among Finnish research publications from 2016-2017 (n=34,507 research articles). As part of the analysis, the authors identified hybrid OA articles through a dedicated OA status metadata field that has been mandatory to report in Finland since 2016 and found a 7% share of hybrid OA. However, the drawbacks of Kirkman's (2018) and Pölönen et al.'s (2020) approaches are that such data sources might not be openly available or, regarding CRIS data, might not be available at all since not all countries maintain comprehensive institutional publication data.

Other studies assessed the share of (hybrid) OA for the entire corpus of scholarly research articles. For instance, using research articles with an assigned DOI and indexed in the Web of Science Core Collection (n=2,610,305) as a benchmark, @Mart_n_Mart_n_2018 assessed how many research articles from 2009 and 2014 are freely available through Google Scholar. After identifying OA articles through licensing information from Crossref, the authors estimated the share of hybrid OA at around 0.5% in 2009 and 1.5% in 2014. In a similar vein, @Piwowar_2018 drew on Crossref to identify content licenses. The authors followed a data-driven approach based on Unpaywall data, which is openly available and contained 66 million Crossref DOIs at the time, to determine how much of the scholarly literature is OA. Based on a random sample of 100,000 Crossref DOIs, the authors found 3.6% of hybrid OA, while pointing towards a growing trend in recent years that recorded a hybrid OA share of 9.4% of all articles published in 2015.

Since its launch, Unpaywall has become widely used as an OA evidence source in bibliometric research and rankings[@Robinson_Garcia_2020; @Huang_2020]. However, results cannot easily be compared with each other over time because of Unpaywall's ongoing changes due to non-standardized data to discover hybrid OA. These changes mainly reflect the growing albeit incomprehensive license information in Crossref metadata. To overcome these shortcomings, the developers extended previous web-scraping approaches to better distinguish between immediate and delayed OA based on licenses at the date of observation @Piwowar_2019. Relatedly, in October 2020, Elsevier's license for delayed OA content was added to Unpaywall's list of license patterns [@Unpaywall_oa_license]. At the journal-level, a more fine-grained identification of full OA journals using OA article information was developed to better distinguish them from subscription journals including hybrid journals [@Unpaywall_types]. With its servie unsub, Unpaywall recently launched a paid service, where libraries can compare hybrid OA uptake rates with internal financial and usage data to assess the value of journal subscriptions [@Chawla_2020].

@@

Some countries like Sweden and the United Kingdom (UK) have negotiated national publishing agreements that allow authors to publish hybrid OA at a discounted fee or for free. These agreements can serve as the basis for studies on the uptake of hybrid OA or cost-benefit analyses (cf. @Olsson1271866; @Lawson_2018) but are often limited in scope as they tend to focus on a single country and rely on data that seldom is publicly available. Hence, while being valuable for specific use-cases, such studies can neither provide a comprehensive understanding of how the scholarly publishing system functions beyond national boundaries nor facilitate collective sharing of financial information about hybrid OA.

Pinfield et al.'s (2016) analysis of APC payment records provided by 23 UK higher education institutions revealed a sharp increase in central payments from 2007-2013, which was largely attributed to the introduction of block grants by Research Councils UK (RCUK) and non-compliance sanctions by the Wellcome Trust. Moreover, the study showed that OA fees were paid almost exclusively through RCUK and Wellcome block grants (92%), and only a small number of APCs were paid through internal funding (7%). In contrast, a recent Springer Nature survey found that authors draw on a range of funding sources to cover OA fees, such as dedicated institutional OA funds, block grants, OA agreements, or research grants [@Monaghan_2020]. Most hybrid OA authors were supported through dedicated institutional OA funds (43%, excluding block grants) and OA agreements with Springer Nature (34%). The differences between these two studies may be due to changes in policy and funding arrangements—in particular, the introduction of OA agreements—since @Pinfield_2016, or due to Monaghan et al.'s (2020) more regionally diverse sample, indicating regional differences in APC funding.

Regional and policy differences in APC payments also came to light in a study by @Jahn_2016 that analyzed APC records from 30 German higher education and research institutions crowd-sourced by the Open APC initiative, the Austrian Science Fund (FWF), Jisc, and the Wellcome Trust. In particular, the study revealed large differences in the amount of hybrid OA funded from 2014-2015. Whilst hybrid OA accounted for less than 1% of APCs paid by German institutions (23 of 3,846), the three non-German research funders recorded a hybrid share of75% (11,533 of 15,779). According to @Jahn_2016, this could point towards differences in science policy, such as hybrid OA being supported by the three non-German research funders but not by Germany's largest national funder (DFG). Another possibility is that German hybrid OA fees were paid from budgets not reported to the Open APC initiative, from where the authors acquired the German APC data. Among these unreported funds are research grants and research unit budgets, which author surveys identified  as funding sources for hybrid OA fees [@Monaghan_2020; @Graaf_2017]. As such, Jahn and Tullney's (2016) findings could reflect the complexities and potential limitations of institutional OA spending data that Pinfield et al. (2016) and Monaghan et al. (2020) attributed to incomplete or missing records.

As this brief review has shown, the most comprehensive studies approach hybrid OA funding from an institutional perspective that focuses on research institutions or funders and draws on internal records or crowd-sourced data like from the Open APC initiative. The advantages of Open APC data are that they reflect the actual APCs paid rather than list prices; they are not limited to certain publishers or journals; they are provided in a standardized format regardless of the publisher. On the other hand, the drawbacks include the limited number of participating institutions that are almost exclusively based in Europe and the dependency on participating organizations for data integrity [@Pieper_2018]. Moreover, since Open APC data does not include waivers or free APCs, it cannot present a complete picture of all publishing activities and expenses. This, however, would pose less of an issue when analyzing just hybrid OA.

## The role of publisher-provided data

Uptake and financial studies have OA policy implications for hybrid OA. Most prominently, the cOAlition S, an international network of research funders, released its widely discussed Plan S. In September 2018, the cOAlition S announced that from 2021 onwards, signatories would no longer fund articles in hybrid OA journals. This principle includes so-called "mirror" and sister journals--hybrid journals with fully OA counterparts or split into their fully OA and subscription parts—that have launched since Plan S was announced but makes an exception for hybrid journals with transformative agreements [@Plan_s; @Frantsv_g_2019; @McNutt_2019; @Harrison_2019]. The Plan S refers to the Efficiency and Standards for Article Charges (ESAC) guidelines [@Geschuhn_2017] as critical component of transformative agreements, emphasizing the need for publisher-provided metadata to address the lack of transparency surrounding hybrid OA. However, studies confirm that publisher with such agreements lack essential data about hybrid OA uptake and invoicing, or are not able or willing to provide such information publicly and instantaneously [@Monaghan_2020; @Marques_2019], which, in turn, not just limit the monitoring of hybrid OA, but also research studies on its size and financial implications.

Here, we focus on Elsevier. The publisher provides article-level metadata to Crossref, including license information, which can be used to identify OA articles, and links to HTML and XML full-text versions for text and data mining purposes. Surprisingly, besides the full-text article, the XML files also contain invoicing information and the article's OA status. Table \ref{tab:els_xml} summarizes the key metadata elements, which comprise the type of entity to which the APC was invoiced (`openaccessSponsorType`) and the name of that entity if available (`openaccessSponsorName`). Moreover, the article-level metadata indicates if OA was provided through Elsevier's Open Archive program (`openArchiveArticle`), which enables free access to participating subscription journals after an embargo period (i.e., delayed OA; @Els_Archive). As we have been unable to identify similar efforts by other large publishers (e.g., Springer Nature), this increased transparency through openly available bibliometric and financial data at the article level seems unique to Elsevier.

\begin{table}[H]
\caption{\label{tab:els_xml} Overview of hybrid open access (OA) uptake across Elsevier hybrid journals with at least one OA article 2015-2019}
\centering
\begin{tabular}[t]{lp{10cm}}
\toprule
XML Node & Description\\
\midrule
\texttt{openaccessArticle} & Indicates if the article is open
access. \\
\texttt{openaccessSponsorType} & Invoice recipient type.\\
\texttt{openaccessSponsorName} & Entity that received Elsevier's APC
invoice.\\
\texttt{openArchiveArticle} & Indicates if open access was provided through
Elsevier's Open Archive programme.\tabularnewline \\
\bottomrule
\end{tabular}
\end{table}

In this study, we utilize Elsevier's publicly available metadata to analyze the uptake and invoicing of their hybrid OA publishing option. In particular, we use this novel methodology to answer the following questions:

- What was the uptake of Elsevier's hybrid OA publishing option between 2015 and 2019?

- Through which channels were hybrid APCs invoiced, and who were the recipients?

@@

For this study, we collected data relating to Elsevier's hybrid OA option by drawing on multiple freely available data sources. We identified Elsevier hybrid journals through Elsevier’s APC list and supplemented our sample with Crossref metadata and text-mined invoicing data to investigate the financing of immediate OA articles provided under a Creative Commons license in subscription-based journals. Figure \ref{fig:workflow} visualizes the automated workflow we used to collect data from Elsevier and Crossref.

@@

\caption{Data Analytics Workflow to obtain article-level OA and invoicing data using freely available data sources provided by Elsevier and Crossref.
}\label{fig:workflow}
\end{figure}

First, we obtained hybrid journals through an Elsevier APC price list from May 2020 provided as captured by @Matthias_2019. We only included hybrid journals that were published under the hybrid model throughout the entire period of our analysis. Hence, we excluded journals that flipped from OA to hybrid and from hybrid to OA [@Matthias_2020]. This approach allowed us to ensure that the article-level differences we observed were not due changes at the journal-level (i.e., increases or decreases in number). Overall, we identified 1,970 unique hybrid journals.

Next, we used an openly available Crossref database snapshot [@Crossref_2020], which contains all Crossref records registered until March 2020, to calculate the journals' combined article volume for the five-year period from 2015-2019. We only included articles published in regular issues aside from supplements containing conference contributions like meeting abstracts, indicated by non-numeric pagination. We excluded non-scholarly journal content, such as the table of contents and lists of reviewers, following Unpaywall's paratext recognition approach [@Unpaywall_para], which we expanded to include patterns indicating corrections. Furthermore, we categorized the articles by subject according to the All Science Journal Classification code (ASJC) and added citation indicators from the Scopus Source Title list (version June 2020) based on the journal they were published in.

Crossref metadata also contains license information that indicates if and under which terms the publisher has made articles openly available on their website [@Hendricks_2020]. We identified OA articles through Creative Commons (CC) licenses and then downloaded the XML version of all CC-licensed articles published in a hybrid journal. From the XML files, we determined whether or not the articles were immediate OA using the XML node `openArchiveArticle` and measured the uptake of hybrid OA through three bibliometric indicators [@Nelson_2017; @Laakso_2016]-—the number of hybrid journals with at least one OA article, the number of hybrid OA articles, and the number of hybrid OA articles relative to the number of closed-accessed articles in hybrid journals with at least one OA article.

Moreover, we obtained APC invoicing metadata from the XML (see Table 1), which enabled us to identify the invoice channels and recipients of hybrid OA APCs. Based on the `openaccessSponsorType` node in the XML files, we distinguish between four invoice channels, including invoices billed to the corresponding author, as part of publishing agreements with funding bodies (hereinafter "agreements"; cf. @Els_Agreements), fee waivers (e.g., in "cases of genuine need" or due to society or university sponsorships, cf. ref), and other types not specified by Elsevier. Finally, when hybrid APCs were invoiced as part of OA agreements, we identified invoice recipients through the `openaccessSponsorName` node.

We manually classified invoice recipients based on their institutional sectors, countries, and primary research areas. Following the OECD's Frascati Manual [@Frascati, p. 91], we coded for four sectors, business enterprise, government, higher education, and private non-profit. The government sector consists of federal, state, and municipal governments, plus non-profit institutions controlled by government units [@Frascati, p. 100]. The higher education sector comprises institutions that perform research and development and provide formal tertiary education services, such as universities or research institutes [@Frascati, p. 101]. Legal or social entities that produce goods and services without generating profit, such as learned societies or charities, constitute the private non-profit sector [@Frascati, p. 288]. Due to the low article volume, we combined invoice recipients Elsevier listed as "authors" and "third-party sponsor", and those we had coded as business enterprises [@Frascati, p. 97], into "Others".  Moreover, we categorized invoice recipients according to the country or countries representing their scope of funding and based on the following primary research areas health sciences, life sciences, physical sciences and mathematics, social sciences and humanities, unknown, and broad (i.e., multiple research areas).

It is important to note that we could not find any official comprehensive documentation of Elsevier's XML schema , at the time of writing. However, our conceptualization of the variables and interpretation of the data was informed by a comprehensive review of Elsevier's websites about OA, including a list of funding agreements and details about Elsevier's Open Archive program,and Elsevier’s article submission process in the context of dedicated OA agreements [@lund; @vsnu; @Els_videos]. For validation, we matched the invoicing data provided by Elsevier to data from  the Open APC initiative.

Throughout the mostly automated data gathering and analysis process, we used tools from the Tidyverse [@tidyverse] for the R programming language [@r]. To allow for efficient data manipulation and retrieval, we imported the Crossref dump to Google BigQuery, applying the rcrossref [@rcrossref] parsers to extract relevant metadata fields. We interfaced Google BigQuery with SQL with the bigrquery client [@bigrquery]. We used crminer [@crminer] to obtain the XML files from Elsevier and the xml2 package [@xml2] to parse relevant metadata. To ensure our results are reproducible, we provided all code for obtaining the data and performing the analyses with detailed documentation as RMarkdown notebooks on GitHub (<https://github.com/njahn82/elsevier_hybrid_invoicing>).

@@

This section first presents the results of our analysis of hybrid OA uptake in Elsevier's journal portfolio with a view to licensing, disciplinary differences, and citation impact. Then, we present a descriptive analysis of Elsevier's invoicing data, highlighting licensing and disciplinary differences for different invoicing channels, institutional sectors, and individual invoice recipients.

@@

Finally, OA policies in favor of hybrid OA in combination with invoicing agreements with Elsevier might be a driver. For instance, we found that nearly half of articles in arts and humanities were facilitated through agreements, which was above-average. This is in line with previous research, suggesting that OA policies and funder mandates have a stronger effect on OA uptake than disciplinary publication cultures [@Huang_2020; @Larivi_re_2018].

>>>>

The rise of open access (OA) has added complexity to
scholarly publishing, particularly concerning transparency of economic
dimensions. Financial transparency in journal publishing has long been
lacking because Big Deal subscription contracts between academic
institutions and large academic publishers usually include
confidentiality clauses that prohibit publicly sharing
the scope or pricing of such agreements [@Frazier_2001; @Bergstrom_2014; @Larivi_re_2015]. Further, the distributed payment processes
to access and publish scholarly articles add to the opacity of
the financial flows of scholarly publishing [@Lawson_2016] as they can
involve several actors simultaneously, including authors, research
institutions, libraries, and funders.

The problem with this lack of transparency becomes more apparent in the
case of hybrid OA. In this OA business model, individual articles can be
made openly available by paying an article processing charge
(APC), while the journal as a whole remains subscription-access. Hence,
hybrid OA was originally introduced as a low-risk transitional model
that allows journals to gradually convert to full OA and reduce
subscription costs as the OA uptake increases [@Prosser_2003]. Many
large subscription publishers have since incorporated hybrid OA, but there have been concerns that the hybrid model allows for double dipping---receiving two payments for one article, the APC and
subscription fees [@Shieber_2009; @Prosser_2015]. While publishers
assure they adjust their pricing
[@Mittermaier_2015], without transparency around the uptake of hybrid OA
and both revenue streams, such claims are impossible to evaluate. This is especially concerning because hybrid OA has recently gained popularity after a decade of only slow uptake [@Bj_rk_2012]. The main drivers of this development have been research institutions and funders that implemented  OA policies, allocated OA funds, and established
formal agreements with publishers that allow
affiliated authors to publish OA free-of-charge [@Laakso_2016].

The recent introduction of transformative agreements, an evolving
concept describing contracts that shift library spending from subscriptions to OA
[@Borrego_2020; @Hinchliffe_2019], might perpetuate the lack of transparency because pricing is often based on traditional subscription costs. Hence, the demand for
publisher-provided data has increased. More
transparency about hybrid OA uptake and funding could  facilitate the assessment and adjustment of publisher contracts [@Schimmer_2015], enhance OA mandate compliance monitoring, and avoid double-dipping [@Larivi_re_2018]. However, previous studies have noted an absence of transparency in hybrid OA publishing and a considerable lack of
publicly available and standardized data [@Laakso_2016; @Lawson_2015; @Pinfield_2016].

This paper investigates to what extent publisher-provided
data can increase the transparency around hybrid OA by taking Elsevier,
the largest scholarly journal publisher [@Larivi_re_2015], as the starting point for an empirical
analysis of hybrid OA uptake and invoicing. Our approach is based on
openly available data sources and combines scholarly metadata
from Crossref, a DOI registration agency, with
 APC invoicing data from Elsevier. The
compiled data enable us to analyze the 
hybrid OA uptake in Elsevier journals between  2015 and 2019 by license and subject, and to compare it with
Elsevier's full and delayed OA program. Further, we determine whether
hybrid APCs were waived or charged to authors or as part of publishing agreements. In the latter case, we 
examine the invoiced academic consortia and research funders. As such, our findings have implications for research and OA
policy and highlight the potential of publisher-provided data for
large-scale, instantaneous analysis of hybrid OA uptake and invoicing.

@@

@@

In the first article-level study on hybrid OA, @Laakso_2016
reported that between 2007 and 2013, the five largest
publishers--Elsevier, Sage, Springer, Taylor \& Francis, and
Wiley--recorded growing numbers of hybrid journals that coincided with a
twenty-fold increase in hybrid OA (from 666 to 13,994). As a
result, hybrid journals with at least one OA article more than doubled
in number (1,082 in 2009, 2,714 in 2013) but fell by 31\% relative to
the total number of hybrid journals. Laakso and Björk's
(2016) exploratory approach yielded a comprehensive overview tracking the growth of hybrid
OA over time and across multiple publishers, but due to the amount of
manual data collection and cleaning it
is unsuitable for repeated use. Minimizing such manual tasks,
Nelson \& Eggett (2017) narrowed their focus to one publisher--the
American Chemical Society (ACS)--and requested a list of all hybrid OA
articles published from 2006-2011 from the ACS. To assess the
hybrid OA uptake, the authors aggregated and compared the number of OA
(n=814) and subscription articles (n=27,621), finding that over the five
years, 2.9\% of ACS articles were published as hybrid OA.

Contrary to the previous two studies, @Kirkman_2018 took a research
funder as a starting point, reporting a 26\% share of hybrid OA for
articles funded by the Australian National Health and Medical Research
Council from 2012-2014 (816 of 3,190). Kirkman collected research
articles using the Funding Agency and Funding Text search fields on Web
of Science, and classified articles as OA when they were freely
available on the publisher website, and further distinguished hybrid OA
articles based on journal-level information from the publisher website,
such as hybrid OA journal and APC lists. Another approach is presented
by @P_l_nen_2020, who used current research information system
(CRIS) data---institutional publication data---to study the extent of OA
among Finnish research publications from 2016-2017 (n=34,507 research
articles). As part of the analysis, the authors identified hybrid OA
articles through a dedicated OA status metadata field that has been
mandatory to report in Finland since 2016 and found a 7\% share of
hybrid OA. However, the drawbacks of Kirkman's (2018) and Pölönen et
al.'s (2020) approaches are that such data sources might not be openly
available or, regarding CRIS data, might not be available at all because
not all countries maintain comprehensive institutional publication data.

Other studies assessed the share of (hybrid) OA for the entire corpus of research articles. Using DOI-assigned research articles indexed in the Web of Science Core Collection
 as a benchmark, @Mart_n_Mart_n_2018 investigated how
many research articles from 2009 and 2014 (n=2,610,305 in total) are freely available through
Google Scholar. After identifying OA articles through licensing
information from Crossref, the authors estimated the share of hybrid OA
at around 0.5\% in 2009 and 1.5\% in 2014. Similarly, in a study assessing the OA level among all scholarly articles and that experienced by Unpaywall users, Piwowar et
al. (2018) obtained license information through Crossref and web-scraping. The study found 3.6\% of 100,000 random Crossref DOIs were hybrid OA
and pointed towards a growing trend in recent years that recorded a
hybrid OA share of 9.4\% of all articles published in 2015.

Since its launch, Unpaywall has become widely used in bibliometric research and rankings [@Robinson_Garcia_2020; @Huang_2020]. However, the lack of standardized and comprehensive publisher-provided data has required several updates to Unpaywall to improve hybrid OA identification and differentiation from  delayed OA [@Piwowar_2019; @Unpaywall_oa_license; @Unpaywall_types], illustrating ongoing challenges in tracking and comparing hybrid OA prevalence over time. After changes in Unpaywall's approach to detect hybrid OA, Piwowar et al. (2019) estimated a hybrid OA share of 4\% amongst all articles published by July 2019.

@@

Pinfield et al.'s (2016) analysis of APC payment records provided by 23
United Kingdom (UK) higher education institutions revealed a sharp increase in central
payments from 2007-2013, which was largely attributed to the
introduction of block grants by Research Councils UK (RCUK) and
non-compliance sanctions by the Wellcome Trust. Moreover, the study
showed that OA fees were paid almost exclusively through block grants (92\%), and only a small number of APCs were paid
through internal funding (7\%). In contrast, a recent Springer Nature
survey found that authors draw on a range of funding sources to cover OA
fees, such as dedicated institutional OA funds, block grants, OA
agreements, or research grants (Monaghan et al., 2020). Most hybrid OA
authors were supported through dedicated institutional OA funds (43\%,
excluding block grants) and OA agreements with Springer Nature (34\%).
The differences between these two studies might be due to Monaghan et al.'s
(2020) focus on Springer Nature articles or due to} different policy and funding arrangements---considering the introduction of OA
agreements since  @Pinfield_2016 and Monaghan et al.'s
(2020) more regionally diverse sample.

Regional and policy differences in APC payments also came to light in a
study by @Jahn_2016 that analyzed APC records from 30 German
higher education and research institutions, the Austrian Science Fund (FWF), Jisc, and the Wellcome
Trust. In particular, the study revealed large differences in the amount
of hybrid OA funded from 2014-2015. Whilst hybrid OA accounted for less
than 1\% of APCs paid by German institutions (23 of 3,846), the three
non-German research funders recorded a hybrid share of 75\% (11,533 of
15,779). According to @Jahn_2016, this could point towards
differences in science policy, such as hybrid OA being supported by the
three non-German research funders but not by Germany's largest national
funder, the German Research Foundation (DFG). Another possibility is that German hybrid OA fees were
paid from budgets not reported to the Open APC initiative, a crowd-sourcing effort [@Pieper_2018] from where
the authors acquired data. Among these unreported funds
are research grants and research unit budgets, which author surveys
identified as APC funding sources [@Monaghan_2020; @Graaf_2017]. As such, Jahn and Tullney's (2016) findings could reflect
the complexities and potential limitations of institutional OA spending
data that @Pinfield_2016 and @Monaghan_2020 attributed
to incomplete or missing records.

In recent years, national consortia in Europe have negotiated publishing agreements covering hybrid OA fees for affiliated authors (Borrego et al., 2020). While these have improved invoicing workflows, internal assessments of hybrid OA uptake and invoicing remain challenging because transparent and comparative data have largely remained absent [@Marques_2020]. Such publicly available information might be scarce because publishers and institutions lack or withhold such data, or due to confidentiality clauses [@Marques_2020; @Monaghan_2020]. Several European consortia have collaborated through the Efficiency and Standards for Article Charges (ESAC) Initiative and Knowledge Exchange to standardize metadata requirements across publishers as great variation has been found between different countries and publishers [@Marques_2019].

## Research questions and aims

In this paper, we focus on Elsevier, a prominent example in recent hybrid OA uptake and financial studies [@Pinfield_2016; @@Laakso_2016]. Elsevier's OA portfolio presents the challenges in examining hybrid OA described above. For instance, distinguishing between different OA types. Elsevier supports delayed [@Els_Archive], hybrid, and full OA, including so-called mirror journals---full OA counterparts of hybrid journals addressing OA policies opposed to hybrid OA [@Harrison_2019]. Further, Elsevier processes APC invoices through various channels, such as agreements with research funders and library consortia or, in the absence thereof, the authors (Elsevier,
n.d.-a, n.d.-c). Surprisingly, we found Elsevier article-level metadata embedded in XML full-texts indicating the articles' OA status and invoicing.  Leveraging this publicly available data, we 
address the lack of transparency around hybrid OA noted in previous studies.
In particular, we use this novel approach to answer the following
questions:

* What was the uptake of Elsevier's hybrid OA publishing option between 2015 and 2019?
* Through which channels were hybrid APCs invoiced, and who were the recipients?

@@

For this study, we collected data relating to Elsevier's hybrid OA
option by drawing on multiple freely available data sources. We
identified Elsevier hybrid journals through Elsevier's APC list and
supplemented our sample with Crossref metadata and text-mined invoicing
data to investigate the invoicing of immediate OA articles provided
under a Creative Commons (CC) license in subscription-based journals. Figure
\ref{fig:workflow} visualizes the automated workflow we used to collect
data from Elsevier and Crossref.

@@

\caption{Data Collection Workflow to Obtain Article-Level OA and Invoicing Data.
}\label{fig:workflow}
\end{figure}

First, we identified hybrid journals through an Elsevier APC price list
from May 2020 provided by @Matthias_2020. We excluded journals that transitioned from hybrid to full OA and
reverse-flip journals that flipped 
from full to hybrid OA [@Matthias_2019]. Overall, we identified 1,970 unique hybrid journals that published between 2015 and 2019.

Next, we used an openly available Crossref database snapshot [@Crossref_2020], which contains all Crossref records registered until March 2020,
to calculate the journals' combined article volume for the five-year
period from 2015-2019. We only included articles published in regular
issues aside from supplements containing conference contributions like
meeting abstracts, indicated by non-numeric pagination. We excluded
non-scholarly journal content, such as the table of contents, following Unpaywall's paratext recognition approach
 [@Unpaywall_para], which we expanded to include patterns indicating
corrections. Furthermore, we categorized the articles by subject
according to the All Science Journal Classification code (ASJC) and
added citation indicators from the Scopus Source Title list (version
June 2020) based on the journal they were published in.

We identified OA articles
through CC licenses in Crossref metadata records (Hendricks et al., 2020) and then downloaded the XML
version of all CC-licensed articles published in a hybrid journal. From
the XML files, we obtained the articles' OA status and invoicing information (see Table \ref{tab:els_xml}). We determined whether articles were immediate or delayed
OA using the XML node \texttt{openArchiveArticle}, which indicates if an article was made freely available under Elsevier's Open Archive program, and measured the
uptake of hybrid OA. Moreover, we obtained the invoice channels and recipients of hybrid OA APCs. Using the \texttt{openaccessSponsorType} node, we distinguished between four invoice channels, including
invoices billed to authors, as part of publishing
agreements with funding bodies (cf. @Els_Agreements), exempted through fee waivers (e.g., in ``cases of genuine need'' or due to
society or university sponsorships, cf.~Elsevier,
n.d.-c), and other types not
specified by Elsevier. Finally, when hybrid APCs were invoiced as part
of agreements, we identified invoice recipients through the
\texttt{openaccessSponsorName} node. It is worth noting that while this article focuses on OA articles published in hybrid journals, the same data is also available for articles published in full OA journals. Elsevier does not provide the invoiced APC in either case.

\begin{table}[H]
\caption{\label{tab:els_xml}Metadata in Elsevier XML Full-Texts.}
\centering
\begin{tabular}[t]{lp{10cm}}
\toprule
XML node & Description\\
\midrule
\texttt{openaccessArticle} & Was the article open
access? \\
\texttt{openaccessSponsorType} & Invoice channel.\\
\texttt{openaccessSponsorName} & Invoice recipient.\\
\texttt{openArchiveArticle} & Was open access provided through
Elsevier's Open Archive program?\tabularnewline \\
\bottomrule
\end{tabular}
\end{table}

We manually classified invoice recipients based on their institutional
sectors, countries, and primary research areas. Following the OECD's
Frascati Manual [@Frascati, p. 91], we coded for four sectors: business
enterprise, government, higher education, and private non-profit. 
Due to the low article volume,
we combined the business enterprise sector with invoice recipients Elsevier listed as "authors" and
"third-party sponsor"
 into "Others". Moreover, we categorized invoice
recipients according to the countries representing their
scope of funding and based on the following primary research areas
health sciences, life sciences, physical sciences and mathematics,
social sciences and humanities, broad (i.e., multiple
research areas), and unknown. Further, we compared Elsevier's invoicing data
with institutional spending data from the Open APC initiative.

Throughout this mostly automated data gathering and analysis process, we
used tools from the Tidyverse [@tidyverse] for the R
programming language [@r]. To allow for efficient data
manipulation and retrieval, we imported the Crossref dump to Google
BigQuery, applying the rcrossref [@rcrossref] parsers to
extract relevant metadata fields. We used crminer
[@crminer] to obtain the XML-full texts from Elsevier.

@@

This section first presents the results of our analysis of hybrid OA
uptake in Elsevier's journal portfolio with a view to licensing,
disciplinary differences, and citation impact. Then, we present a
descriptive analysis of Elsevier's invoicing data, highlighting
licensing and disciplinary differences for invoicing channels and invoice recipients.

@@

Finally, OA policies in favor of hybrid OA in combination with invoicing agreements with Elsevier might be a driver. For instance, we found that nearly half of articles in arts and humanities were facilitated through agreements, which was above-average. This is in line with previous research, suggesting that OA policies and funder mandates have a stronger effect on OA uptake than disciplinary publication cultures [@Huang_2020; @Larivi_re_2018].  While these aspects are important drivers of both full and hybrid OA, faculty members generally prioritize journal prestige and the journal impact factor [@Niles_2020] in publishing decisions, which, in combination with the availability of OA funding, might encourage publishing hybrid OA [@van_der_graaf_maurits_2017].
